Rank verification for exponential families

### 2016-04-19 verifying-winner.tex

\documentclass[11pt]{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{caption}
\usepackage{cleveref}
\usepackage{enumitem}
\usepackage{graphicx}
\usepackage[totalwidth=480pt, totalheight=680pt]{geometry}
\usepackage{mathrsfs}
\usepackage{tikz}
\usetikzlibrary{calc}
\usepackage{xcolor}
\definecolor{mid_green}{HTML}{99d8c9}

@@

\begin{abstract}
This paper deals with the selective inference problem encountered in rank verification --- verifying the order of the natural parameters in exponential families. We constructed a selective level-$\alpha$ test for verifying the largest natural parameter, which provides familywise error rate (FWER) and false discovery rate (FDR) control as we extend it into a step-down procedure. We also provide selective confidence bound for the difference between the largest and the second largest natural parameters.
\end{abstract}

@@

We often make observations that form a joint distribution. While it might be relatively obvious in the independent case, the observations can well be dependent, complicating comparisons of parameters. However, comparison of these parameters can produce useful inferences.

A classic question is on how to statistically verify empirical rankings based on simple exponential family models. As motivation, consider a 2016 Quinnipiac University poll of $890$ Iowa Republicans Poll. \cite{quinnipiac} The poll found that Trump was in the lead with $31\%$ of the vote, Cruz is the first runner-up with $24\%$, followed by Rubio at $17\%$. The other ten candidates, including ``Don't know'', trailed behind. \Cref{tbl:poll} shows the results of the poll. Note that we have made a simplifying assumption that poll represents a random sample of Iowa Republicans --- i.e.\ that the data are multinomial, with parameter $\left(\pi_{\text{Trump}}, \pi_{\text{Cruz}}, \pi_{\text{Rubio}}, \ldots\right)$ representing their support in the Iowa Republican population. In reality, Quinnipiac University Poll Institute has processed the data to make the sample more representative of the population of likely voters.

@@

Having observed that Trump received the most votes in this poll, we want to ask whether he is actually in the lead in the population of all Iowa Republicans. Intuitively, we would compare Trump to Cruz, as well as the other candidate. This intuitive method however has a couple of issues.

@@

\item Comparing Trump to all other candidates involves multiple comparisons. Without considering the structure of this problem, common methods that perform multiple comparisons may largely decrease the test power.

@@

Next we condition on the sufficient statistics that are multiplied to the the nuisance parameter, in other words, we consider the conditional law
$$\mathcal{L}_{\pi_1 = \pi_j}\left(\frac{X_1 - X_j}{2} \middle| X_2, \ldots, \frac{X_1 + X_j}{2}, \ldots, X_n, A_1\right),$$
or equivalently, the conditional law
$$\mathcal{L}_{\pi_1 = \pi_j}\left(X_1 \middle| X_2, \ldots, \frac{X_1 + X_j}{2}, \ldots, X_n, A_1\right),$$
which is the distribution $\text{Binomial}\left(X_1 + X_j, \frac{1}{2}\right)$ truncated by the event $A_1$. The one-sided $p$-value $p_{1j}$ is thus given as
\begin{align*}
p_{ij}\left(x_1, \ldots, x_n\right) & = \PP\left(X_1 \ge x_1 \middle| X_2 = x_2, \ldots, \frac{X_1 + X_j}{2} = \frac{x_1 + x_j}{2}, \ldots, X_n = x_n, A_1\right) \\
& = \frac{\sum_{t = x_1}^{x_1 + x_j} \binom{x_1 + x_j}{t} \left(\frac{1}{2}\right)^{x_1 + x_j}}{\sum_{t = \left(x_1 + x_j\right)/2}^{x_1 + x_j} \binom{x_1 + x_j}{t} \left(\frac{1}{2}\right)^{x_1 + x_j}} \\
& = \frac{\sum_{t = x_1}^{x_1 + x_j} \binom{x_1 + x_j}{t}}{\sum_{t = \left(x_1 + x_j\right)/2}^{x_1 + x_j} \binom{x_1 + x_j}{t}},
\end{align*}
and to account for the tie-breaking we adopted earlier, we
\begin{itemize}
\item multiply the first term in the numerator and the denominator by $1/k$ where $k$ is the number of observations tied for largest; otherwise,
\item multiply the first term in the denominator by $1/2$ if $X_1$ and $X_j$ has the same parity, as the sum starts $X_1 = X_j = \frac{x_1 + x_j}{2}$.
\end{itemize}
Noticing that if $X_1, \ldots, X_j$ all tied the $p$-value is $1$, and the denominator is really $1/2$ otherwise \WFcomment{I don't think that's true for $j \neq 2$, is it?} \KHcomment{The denominator should always be half of a binomial distribution I think}, we can write
$$p_{1j} = 1 \wedge 2 \sum_{t=X_1}^{X_1 + X_j} \binom{X_i + X_j}{t}.$$

@@

\Cref{fig:power} gives the power curves for $\text{Multinomial}\left(m, \pi\right)$ and
$$\pi \propto \left(e^\delta, 1, \ldots, 1\right),$$
for various combinations of $m$ and $n$. For Gupta and Nagel, we use $\alpha = 0.05$; but to account for the extra factor of $\frac{n-1}{n}$ in \Cref{eqn:selective_vs_marginal}, we will apply the selective procedure for with $\frac{n}{n-1} \alpha$ such that the marginal type I error rate of both procedures are controlled at $\alpha$. \WFcomment{Can we justify this? Our procedure has error $\alpha$ in the case where all $\pi_j$ are equal.} \KHcomment{The selective error rate is $\alpha$, but Gupta and Nagel really only had marginal error rate. If all $\pi_j$ are equal, the marginal error rate is $\frac{n-1}{n} \alpha$.} Gupta and Nagel coincides with our test at $n = 2$; however as $n$ grows, the selective test shoes significantly more power than Gupta and Nagel's test.

@@

We would like to extend the result from multinomials to exponential families. We henceforth focus on exponential distribution taking the form
\begin{equation}
X \sim \exp\left(\theta_1 X_1 + \cdots + \theta_n X_n\right) g\left(X\right) h\left(\theta\right),
\label{eqn:exp_family}
\end{equation}
defined on the Lebesgue or counting measure. It is reasonable to assume that the family is ``symmetric'', i.e.\ if $\theta_j = \theta_k$, then the probability distribution remains the same when $X_j$ and $X_k$ are swapped. This corresponds to requiring $g\left(X\right)$ to be symmetric. We would also like to introduce a fairly mild additional condition that $g$ is {\em Schur-concave}. But first we recall the notion of {\em majorization}, defined on both $\mathbb{R}^n$ and $\mathbb{Z}^n$. \WFcomment{I think we're implicitly assuming here  that $g$ is a density wrt one of (a) Lebesgue or (b) counting measure. Is that right? And if so, do we need a couple sentences to explain why it works in the counting measure case? That is, we need the atoms of non-zero measure to match up with each other along the rays of interest, right?} \KHcomment{I think we are making that assumption. Added a few more words there but not sure if it is explicit enough.}

\begin{definition}
For two vectors $a$ and $b$ in $\RR^n$, suppose permuting the two vectors in descending order gives
$a_{\left(1\right)} \ge \cdots \ge a_{\left(n\right)}$ and $b_{\left(1\right)} \ge \cdots \ge b_{\left(n\right)}$. We say that $a \succeq b$ ($a$ majorizes $b$) if for $1 \le i < n$,
\begin{align*}
a_{\left(1\right)} + \cdots + a_{\left(i\right)} & \ge b_{\left(1\right)} + \cdots + b_{\left(i\right)} \\
a_{\left(1\right)} + \cdots + a_{\left(n\right)} & = b_{\left(1\right)} + \cdots + b_{\left(n\right)}
\end{align*}
This forms a partial order in $\RR^n$.
\end{definition}

@@

\bibliographystyle{plain}
\bibliography{papers,additional}

>>>>

\documentclass[11pt]{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{caption}
\usepackage{cleveref}
\usepackage{enumitem}
\usepackage{graphicx}
\usepackage[totalwidth=480pt, totalheight=680pt]{geometry}
\usepackage{mathrsfs}
\usepackage{tikz}
\usetikzlibrary{calc}
\usepackage{xcolor}
\usepackage[round]{natbib}
\definecolor{mid_green}{HTML}{99d8c9}

@@

\begin{abstract}
Many statistical experiments involve comparing of multiple population groups. For example, a public opinion poll may ask which of several political candidates commands the most support in a given population, or a clinical trial may compare patient outcomes under several treatment conditions. This article concerns the problem of {\em rank verification} --- post hoc tests for whether the ordering discovered in the data is in fact statistically significant. For exponential family models, we show under mild conditions that an unadjusted two-tailed pairwise $p$-value comparing the ``winning'' candidate to the ``runner-up'' is a valid $p$-value for checking whether the winner is truly the best. We extend this result to provide lower confidence bounds on the gap between the ``winning'' population and the others, and propose a simple step-down algorithm for verifying ranks beyond the first.
\end{abstract}

@@

\subsection{Motivation: Iowa Republican Caucus Poll}

On February 1, 2016, Quinnipiac University released a poll of 890 Iowa Republicans regarding their preferred candidate for the Republican presidential nomination \citep{quinnipiac}. The results, reproduced here in Table~\ref{tbl:poll}, showed Donald Trump in the lead with $31\%$ of the vote, with Ted Cruz in second with $24\%$, Marco Rubio in third with $17\%$, and ten other candidates including ``Don't know'' trailing behind. We make a simplifying assumption that the poll represents a simple random sample of Iowa Republicans, i.e., that the data are a multinomial sample of size $890$ and underlying probabilities $(\pi_{\text{Trump}}, \pi_{\text{Cruz}}, \pi_{\text{Rubio}}, \ldots)$\footnote{The reality is a bit more complicated: before releasing the data, Quinnipiac has processed it to make the reported result more representative of likely caucus-goers.}

@@

Once we see the results of the poll, several salient questions will likely occur to us: Is Trump really winning, and if so by how much? And analogously, is Cruz really in second, is Rubio really in third, and so on?

Note that there is implicitly a problem of multiple comparisons here, since we would be asking a different set of questions if Cruz had led the poll instead. Indeed, the selection issue appears especially pernicious due to the so-called ``winner's curse:'' given that Trump leads the poll, it more likely than not overestimates his support.

Nevertheless, if we blithely ignore the selection issue, we might carry out the following analyses to answer the questions we posed before at significance level $\alpha=0.05$:
\begin{enumerate}
\item {\em Is Trump really winning?} If Trump and Cruz were in fact tied, then Trump's share of their combined 490 votes should be distributed as $\text{Binom}(490, 0.5)$. Because the (two-tailed) $p$-value for this test is $p=0.006$, we can reject the null and declare that, yes, Trump is really winning.
\item {\em By how much?} Using an exact $95\%$ interval for the same binomial model, we conclude Trump has at least $7.5\%$ more support than Cruz (and by extension everyone else).
\item {\em Is Cruz in second, Rubio in third, etc.?} We can next compare Cruz to Rubio just as we compared Trump to Cruz (again rejecting because 214 is significantly more than half of 365), then Rubio to Carson, and so on, as a stepdown procedure. The first four comparisons are all significant at level $0.05$, but Paul and Bush are tied so we stop.
\end{enumerate}

Surprisingly, the three procedures described above are all entirely valid despite the implicit multiple-comparisons issue. The remainder of this article is dedicated to justifying them.

@@

\item Comparing Trump to all other candidates involves multiple comparisons. Without considering the structure of this problem, common methods that perform multiple comparisons may largely decrease the test power.
\end{enumerate}

@@

Next we condition on the sufficient statistics that are multiplied to the the nuisance parameter, in other words, we consider the conditional law
$$\mathcal{L}_{\pi_1 = \pi_j}\left(\frac{X_1 - X_j}{2} \middle| X_2, \ldots, \frac{X_1 + X_j}{2}, \ldots, X_n, A_1\right),$$
or equivalently, the conditional law
$$\mathcal{L}_{\pi_1 = \pi_j}\left(X_1 \middle| X_2, \ldots, \frac{X_1 + X_j}{2}, \ldots, X_n, A_1\right),$$
which is the distribution $\text{Binomial}\left(X_1 + X_j, \frac{1}{2}\right)$ truncated by the event $A_1$. The one-sided $p$-value $p_{1j}$ is thus given as
\begin{align*}
p_{ij}\left(x_1, \ldots, x_n\right) & = \PP\left(X_1 \ge x_1 \middle| X_2 = x_2, \ldots, \frac{X_1 + X_j}{2} = \frac{x_1 + x_j}{2}, \ldots, X_n = x_n, A_1\right) \\
& = \frac{\sum_{t = x_1}^{x_1 + x_j} \binom{x_1 + x_j}{t} \left(\frac{1}{2}\right)^{x_1 + x_j}}{\sum_{t = \left(x_1 + x_j\right)/2}^{x_1 + x_j} \binom{x_1 + x_j}{t} \left(\frac{1}{2}\right)^{x_1 + x_j}} \\
& = \frac{\sum_{t = x_1}^{x_1 + x_j} \binom{x_1 + x_j}{t}}{\sum_{t = \left(x_1 + x_j\right)/2}^{x_1 + x_j} \binom{x_1 + x_j}{t}},
\end{align*}
and to account for the tie-breaking we adopted earlier, we
\begin{itemize}
\item multiply the first term in the numerator and the denominator by $1/k$ where $k$ is the number of observations tied for largest; otherwise,
\item multiply the first term in the denominator by $1/2$ if $X_1$ and $X_j$ has the same parity, as the sum starts $X_1 = X_j = \frac{x_1 + x_j}{2}$.
\end{itemize}
Noticing that if $X_1, \ldots, X_j$ all tied the $p$-value is $1$, and the denominator is really $1/2$ otherwise \WFcomment{I don't think that's true for $j \neq 2$: say $X=(100, 99, 10)$. Then the denominator is }, we can write
$$p_{1j} = 1 \wedge 2 \sum_{t=X_1}^{X_1 + X_j} \binom{X_i + X_j}{t}.$$

@@

\Cref{fig:power} gives the power curves for $\text{Multinomial}\left(m, \pi\right)$ and
$$\pi \propto \left(e^\delta, 1, \ldots, 1\right),$$
for various combinations of $m$ and $n$. For Gupta and Nagel, we use $\alpha = 0.05$; but to account for the extra factor of $\frac{n-1}{n}$ in \Cref{eqn:selective_vs_marginal}, we will apply the selective procedure for with $\frac{n}{n-1} \alpha$ such that the marginal type I error rate of both procedures are controlled at $\alpha$. \WFcomment{Can we justify this? Our procedure has error $\alpha$ in the case where all $\pi_j$ are equal.} \KHcomment{The selective error rate is $\alpha$, but Gupta and Nagel really only had marginal error rate. If all $\pi_j$ are equal, the marginal error rate is $\frac{n-1}{n} \alpha$.} \WFcomment{Yes OK I think you're right. Because for G \& N the ``marginal error rate'' is defined by arbitrarily marking one of the tied populations as the best and then asking whether that marked one gets into the confidence set.} Gupta and Nagel coincides with our test at $n = 2$; however as $n$ grows, the selective test shoes significantly more power than Gupta and Nagel's test.

@@

We would like to extend the result from multinomials to exponential families. We henceforth focus on exponential distribution taking the form
\begin{equation}
X \sim \exp\left(\theta_1 X_1 + \cdots + \theta_n X_n\right) g\left(X\right) h\left(\theta\right),
\label{eqn:exp_family}
\end{equation}
defined on the Lebesgue or counting measure. It is reasonable to assume that the family is ``symmetric'', i.e.\ if $\theta_j = \theta_k$, then the probability distribution remains the same when $X_j$ and $X_k$ are swapped. This corresponds to requiring $g\left(X\right)$ to be symmetric. We would also like to introduce a fairly mild additional condition that $g$ is {\em Schur-concave}. But first we recall the notion of {\em majorization}, defined on both $\mathbb{R}^n$ and $\mathbb{Z}^n$.
\begin{definition}
For two vectors $a$ and $b$ in $\RR^n$, suppose permuting the two vectors in descending order gives
$a_{\left(1\right)} \ge \cdots \ge a_{\left(n\right)}$ and $b_{\left(1\right)} \ge \cdots \ge b_{\left(n\right)}$. We say that $a \succeq b$ ($a$ majorizes $b$) if for $1 \le i < n$,
\begin{align*}
a_{\left(1\right)} + \cdots + a_{\left(i\right)} & \ge b_{\left(1\right)} + \cdots + b_{\left(i\right)} \\
a_{\left(1\right)} + \cdots + a_{\left(n\right)} & = b_{\left(1\right)} + \cdots + b_{\left(n\right)}
\end{align*}
This forms a partial order in $\RR^n$.
\end{definition}

@@

\bibliographystyle{plainnat}
\bibliography{papers,additional}
